Bayesian assessment of heterogeneity in psychological phenomena

### 2023-06-27 ms.Rmd
Clarification
Update draft

---
title: 'Bayesian assessment of heterogeneity in psychological phenomena'
shorttitle: Heterogeneity in psychology
leftheader: Heterogeneity in psychology
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Department of Social Psychology, Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
  - name: Niall Bolger
    affiliation: 2
    address: Department of Psychology, Columbia University
    email: nb2229@columbia.edu
affiliation:
  - id: 1
    institution: Department of Social Psychology, Tilburg University
  - id: 1
    institution: Department of Psychology, Columbia University
abstract: |
  Advances in statistics and data collection methods have brought variability in psychological phenomena to the forefront of theoretical development. Yet, established methods and practices for calculating and reporting heterogeneity are lacking in their treatment of uncertainty regarding estimates that describe the magnitude of heterogeneity in the phenomenon under study. There is then a need for more complete treatments of heterogeneity in psychological phenomena, which we aim to deliver in the current work.
  We provide a walkthrough of probabilistic methods for i. estimating, ii. summarising, and iii. reporting heterogeneity using bayesian hierarchical models. These methods are especially suitable for investigating heterogeneity because they naturally describe uncertainty in all model parameters.
  We hope that the proposed methods help researchers to move from reporting point estimates to drawing inferences about magnitudes and differences in heterogeneity in psychological phenomena across different populations.
keywords: variation, heterogeneity, hierarchical models, mixed models, methodology
authornote: |
  \noindent \textbf{Working paper. Proceed with caution.}
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: references.bib
floatsintext: yes
numbersections: false
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
header-includes:
  - |
    \usepackage{float}
    \setlength{\parskip}{4pt}
output: 
  papaja::apa6_pdf:
    number_sections: true
    keep_tex: false
    highlight: kate
  papaja::apa6_docx: 
    number_sections: true
    keep_tex: false
---

```{r setup, include = FALSE}
# Packages
library(papaja)
library(janitor)
library(scales)
library(ellipse)
library(knitr)
library(lme4)
library(patchwork)
library(latex2exp)
library(ggpp)
library(distributional)
library(ggstance)
library(tidybayes)
library(parameters)
library(brms)
library(tidyverse)

@@

# Plotting options
theme_set(
  theme_linedraw(base_size = 12) +
    theme(
      strip.background = element_rect(fill = "gray90", color = "gray90"),
      strip.text = element_text(color = "black", hjust = 0),
      panel.grid = element_blank()
    )
)

# Model estimation options
dir.create("models", FALSE)
options(mc.cores = 4)
```

@@

# Save variables as numerical & factor (might need for more complex contrast/eff/etc codes)
dat <- read_csv("data/heterogeneity_dataset1_traitvalence.csv") %>% 
  # Cleaning as in Bolger et al 2019
  filter(
    response.keys == "up",
    scale(rt) < 3,
    !(id %in% c(250, 257, 272))
  ) %>% 
  # Some more cleaning
  mutate(
    person = fct_anon(factor(id)),
    valence = factor(valenceE, levels = c(-0.5, 0.5), labels = c("Negative", "Positive")),
    rt = logrt,
    .keep = "none"
  ) %>% 
  select(person, valence, rt) %>% 
  arrange(person)
```

@@

When building and testing theories of psychological phenomena, psychologists have long focused on asking whether an effect exists or not. Yet, establishing that an independent variable affects a dependent variable on average may not be a sufficient description of the phenomenon at hand if the treatment effect is highly variable from one individual to another.

@@

Our aim is to outline methods for studying and communicating heterogeneity in a manner that fully takes uncertainty into account. In the first section, we introduce the example data used throughout the manuscript and review established procedures for estimating and communicating the expected heterogeneity in causal effects in the population.

We then turn to our proposed method of interrogating and communicating causal effect heterogeneity using probabilistic methods: We briefly review key principles of probabilistic inference and how they affect inferences about heterogeneity. We then return to the data example and examine effect heterogeneity from different angles using probabilistic inference.

We further extend the practice and consequent benefits of probabilistic inference for causal effect heterogeneity to comparing heterogeneity between populations and across timepoints in further example analyses.

@@

```{r}
#| label: tab-dat1
dat %>% 
  head() %>% 
  apa_table(
    span_text_columns = FALSE, 
    caption = "First six rows of example dataset 1.", 
    note = "rt is log-transformed."
  )
```

In this section, we replicate @bolgerCausalProcessesPsychology2019's analysis, using their openly available data (https://github.com/kzee/heterogeneityproject). We first cleaned the data as in @bolgerCausalProcessesPsychology2019, leaving `r nrow(dat)` trials from 59 participants that were endorsed as self-relevant (see Table \@ref(tab:tab-dat1)). Then, we estimated the same statistical model with these data:

\begin{equation}
\text{RT}_{ij} \sim \operatorname{Normal}(\eta_{ij}, \sigma^2).
\end{equation}

In words, we model the (log) reaction time of subject $j$ on trial $i$ as a random draw from a normal distribution with mean $\eta$ (*eta*, which can differ between trials $i$ and individuals $j$) and standard deviation $\sigma$ (*sigma*). The lack of subscripts on $\sigma$ indicates that we assume it to be invariant across trials and subjects, but note that this assumption is straightforward to relax. Then, we specify a model of the mean $\eta_{ij}$ such that the regression coefficients capture our substantive questions:

\begin{equation}
\eta_{ij} = \beta_0 + \gamma_{0j} + (\beta_1 + \gamma_{1j})\text{V}_{ij}.
\end{equation}

This equation includes two kinds of parameters: $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V), do not have subscripts. That means that they are not further modelled, and therefore sometimes called "fixed". We contrast coded valence such that negative stimuli are coded as -0.5, and positive stimuli as 0.5. This coding results in an intercept that corresponds to the average reaction time across negative and positive stimuli, and a slope coefficient that reflects the magnitude of the difference in log(rt) between negative and positive words (see Appendix). The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$ are subject-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the average reaction time for subject $j=1$. The $\gamma$ coefficients have subscripts, which in this context indicates that they themselves are further modelled, and therefore sometimes called "random" [@gelmanDataAnalysisUsing2007]. As is standard in multilevel modelling, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

\begin{equation}
\begin{bmatrix} 
\gamma_{0} \\ \gamma_{1}
\end{bmatrix} \sim 
\operatorname{MVN}(
\begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
\begin{bmatrix} 
\tau_{\gamma_{0}} &\rho_{\beta_0\beta_1} \\ 
\rho_{\beta_0\beta_1} &\tau_{\gamma_{1}} 
\end{bmatrix}
).
\end{equation}

In these equations, we assume that the subject-specific deviations are have means of zero, standard deviations $\tau$ (*tau*), and a correlation $\rho$ (*rho*). Note that the $\gamma$s are zero-centered because the means are added to them in the above linear mixed-effects equation for $\eta_{ij}$. What these equations mean substantively is that the extent to which the effect of valence varies around the distribution's mean $\beta_1$ is estimated by the standard deviation $\tau_{\gamma_1}$, and $\rho$ indicates the extent to which individuals who have greater slopes also have greater or smaller average reaction times. With data described in Table \@ref(tab:tab-dat1) loaded to R, we can estimate this model using standard maximum likelihood methods as implemented in the R package lme4 [@batesFittingLinearMixedEffects2015; @rcoreteamLanguageEnvironmentStatistical2022] (see the Appendix for code details.)

@@

```{r fit1-brms}
fit1 <- brm(
  rt ~ 1 + valence + (1 + valence | person),
  data = dat,
  file = "models/brm-fit-1"
)
```

@@

bind_cols(
  tab1_lmer %>% 
    mutate(
      across(where(is.numeric), ~number(., .01)),
      MLE = str_glue("{Coefficient} [{CI_low}, {CI_high}]"),
    ) %>% 
    select(Parameter, MLE),
  tab1_brms %>% 
    mutate(
      across(where(is.numeric), ~number(., .01)),
      Bayes = str_glue("{Mean} [{CI_low}, {CI_high}]"),
    ) %>% 
    select(Bayes)
) %>% 
  mutate(
    Parameter = m1_parnames
  ) %>% 
  apa_table(
    span_text_columns = FALSE,
    caption = "Parameter estimates from Model 1.",
    note = "MLE: Maximum Likelihood Estimate. Numbers in brackets are 95\\% Confidence (MLE) or Credibility (Bayes) Intervals (CIs). CIs for MLE (co)variance parameters are calculated with 100 bootstrap iterations.",
    escape = FALSE
  )
```

We show a traditional summary of the estimated parameters from this model in the MLE column of Table \@ref(tab:tbl-fit-1). The expected effect of positive valence on RT for the average person is approximately `r number(tab1_lmer[2,2,T], .01)` log seconds. Moreover, the expected standard deviation describing this effect's degree of variability in the population is approximately `r number(tab1_lmer[4,2,T], .01)`. These point estimates together define the expected normal distribution of valence effects in the population (see Figure \@ref(fig:hdist-1)).

```{r}
#| label: hdist-1
#| fig.height: 5
#| fig.cap: \textbf{Top}. Expected heterogeneity distribution of valence effects from Example 1. The interval displays the 90\% heterogeneity interval ($HI_{90}$), which captures the central 90\% of values in the distribution. \textbf{Bottom}. Model-estimated person-specific valence effects ($\gamma_1$; dark points) and corresponding coefficients from linear regression models estimated separately for each person (light points).

fit1_coef_lm <- tibble(fit1$data) %>% 
  mutate(person = factor(as.numeric(person))) %>% 
  group_by(person) %>% 
  group_modify(~as_tibble(coef(lm(rt ~ valence, data = .))[2])) %>% 
  ungroup() %>% 
  select(person, Coefficient = value) %>% 
  mutate(model = "lm")
fit1_coef_lmer <- tibble(coef(fit1_lmer)$person)[,2] %>% 
  rownames_to_column("person") %>% 
  select(person, Coefficient = valence1) %>% 
  mutate(model = "lmer")
fit1_coef_brms <- as_tibble(coef(fit1)$person[,,"valence1"]) %>% 
  rownames_to_column("person") %>% 
  select(person, Coefficient = Estimate) %>% 
  mutate(model = "brms")

coefs <- bind_rows(
  fit1_coef_lm,
  fit1_coef_lmer,
  fit1_coef_brms
) %>% 
  pivot_wider(names_from = model, values_from = Coefficient) %>% 
  arrange(lm) %>% 
  mutate(i = 1:n())

(ggplot() +
    coord_cartesian(xlim = c(-0.7, 0.4)) +
    stat_halfeye(
      aes(
        xdist = dist_normal(tab1_brms$Mean[2], tab1_brms$Mean[4]),
      ), 
      .width = c(.90), 
      p_limits = c(.0001, .9999), 
      slab_fill = "grey70",
      slab_color = "grey10",
      slab_linewidth = .4,
      size = 2, justification = .01
    ) +
    stat_halfeye(
      aes(
        xdist = dist_normal(
          mean(fit1_coef_lm$Coefficient), 
          sd(fit1_coef_lm$Coefficient)
        )
      ),
      p_limits = c(.0001, .9999), 
      point_interval = NULL,
      slab_fill = NA,
      slab_color = "black",
      slab_linewidth = .4
    ) +
    scale_y_continuous(
      expand = expansion(c(0.0025, 0.05))
    ) +
    labs(
      y = "Probability",
      x = "Valence effect"
    ) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title.x = element_blank(),
      plot.margin = unit(c(8,8,0,8), "pt")
    )) /
  (coefs %>% 
     ggplot(aes(brms, i)) +
     coord_cartesian(xlim = c(-0.7, 0.4)) +
     scale_y_continuous(
       "Person",
       expand = expansion(0.025)
     ) +
     scale_x_continuous(
       "Valence effect"
     ) +
     geom_segment(
       aes(x = lm, xend = brms, y = i, yend = i),
       linewidth = .1
     ) +
     geom_point(aes(lm), shape = 21, fill = "white") +
     geom_point(shape = 21, fill = "grey70") +
     theme(
       plot.margin = unit(c(0,8,8,8), "pt"),
       axis.text.y = element_blank(),
       axis.ticks.y = element_blank(),
     )
  ) +
  plot_layout(heights = c(3.5, 6.5))
```

```{r hi90, include = FALSE}
hi90_num <- tab1_lmer[2,2,T] + qnorm(c(0.05, .95))*tab1_lmer[4,2,T]
hi90 <- paste(
  number(hi90_num, .01), 
  collapse = ", "
)
hi90_low <- paste(
  number(tab1_lmer[2,3,T] + qnorm(c(0.05, .95))*tab1_lmer[4,3,T], .01), 
  collapse = ", "
)
```

We can use the numbers in Table \@ref(tab:tbl-fit-1) to communicate the degree of heterogeneity in the effect of valence in the population: The standard deviation $\tau_1=$ `r number(tab1_lmer[4,2,T], .01)`. However, that quantity alone is unable to properly characterize the heterogeneity in actionable terms. We were therefore motivated to find alternative concise summaries of the degree of heterogeneity of slopes in the population. To begin with, following @bolgerCausalProcessesPsychology2019, we used the point estimates in Table \@ref(tab:tbl-fit-1) to construct an expected *heterogeneity interval* that describes a range within which a certain percentage of the population's slopes are expected to fall (interval in bottom of Figure \@ref(fig:hdist-1)). Later on, we describe alternative summaries of heterogeneity that address different questions about variability, and afford different ways of communicating and visualizing heterogeneity.

By convention, @bolgerCausalProcessesPsychology2019 and others have focused on the 95% heterogeneity interval ($HI_{95}$), but because there are already confusingly many quantities using the five percent cutoff, in this manuscript we focus on the 90% heterogeneity interval, and reserve 95% to describing uncertainties in estimated quantities. This heterogeneity interval can be calculated by first specifying the desired probability limits. For a 90% interval, we use the .05 and .95 percentiles. Then, we pass those percentiles to the normal quantile function ($q$) using the estimated mean and standard deviation to obtain the interval: $HI_{90} = q([.05, .95], \hat\beta_1, \hat\tau_1)$. Therefore (see Table \@ref(tab:tbl-fit-1)), $HI_{90} =$ $q([.05, .95],$ `r number(tab1_lmer[2,2,T], .01)`, `r number(tab1_lmer[4,2,T], .01)`) = [`r hi90`]. In words, this equation calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates, indicated with hats in the equation: We expect 90% of the effects in the population to fall within this interval.

Critically, while valid, this calculation ignores the uncertainty inherent in the estimated parameters. That is, the uncertainty with which we have estimated the population parameters, indicated by e.g. the confidence intervals in Table \@ref(tab:tbl-fit-1), is not carried forward in calculating $HI_{90}$: We have only presented the expected heterogeneity interval, but not the range of credible values that it might take. For instance, if we re-calculate $HI_{90}$ based on the lowest CI bounds in Table \@ref(tab:tbl-fit-1), the result is a heterogeneity interval that indicates 90% of the population's slopes falling in the [`r hi90_low`] interval. In contrast to the first interval, this interval implies much less heterogeneity and that 90% of the population's slopes are expected to fall below zero.

We have now arrived at the crux of the current work: How should we estimate and communicate causal effect heterogeneity such that the fundamental uncertainty in the estimated quantities (the mean and standard deviation of the distribution of slopes in the population) is retained?

# Interlude on probabilistic estimation in practice {-}

<!-- todo introduce citations and topics as we go; delete this section -->

We propose that probabilistic methods are uniquely applicable to the problem presented above. In what follows, we describe practical calculations and visualizations that make use of bayesian methods in estimating and communicating causal effect heterogeneity. Frequentist methods, such as maximum likelihood above, aim at estimating optimum values of estimated quantities, typically called point estimates. In contrast, probabilistic methods aim at describing distributions of plausible values that could underlie hypothetical data generating processes. In this section, we briefly describe bayesian estimation such that we can meaningfully use it in the remainder of our proposition. For more complete treatments, we refer readers to excellent existing literature on the topic [@gelmanBayesianDataAnalysis2013; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014; @mcelreathStatisticalRethinkingBayesian2020].

Lorem ipsum bayesian propagandum.

>>>>

---
title: 'Bayesian assessment of heterogeneity in psychological phenomena'
shorttitle: Heterogeneity in psychology
leftheader: Heterogeneity in psychology
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Tilburg School of Social and Behavioral Sciences, Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
  - name: Matthew Kay
    affiliation: 2
  - name: Niall Bolger
    affiliation: 3
affiliation:
  - id: 1
    institution: Department of Social Psychology, Tilburg University
  - id: 2
    institution: Computer Science and Communication Studies, Northwestern University
  - id: 3
    institution: Department of Psychology, Columbia University
abstract: |
  Advances in statistics and data collection methods have brought variability in psychological phenomena to the forefront of theoretical development. Yet, established methods and practices for calculating and describing heterogeneity are lacking in how they treat uncertainty regarding heterogeneity in the phenomenon under study. In the current work, we aim to deliver methods and guidelines for more complete treatments of heterogeneity in psychological phenomena.
  We provide a walkthrough of probabilistic methods for i. estimating, ii. summarising, and iii. describing heterogeneity. The methods we propose---bayesian hierarchical models---are especially suitable for investigating heterogeneity because they naturally describe uncertainty in all model parameters.
  We hope that the proposed methods help researchers to move from reporting point estimates to drawing inferences about magnitudes and differences in heterogeneity in psychological phenomena across different populations.
keywords: variation, heterogeneity, hierarchical models, mixed models, methodology
authornote: |
  \noindent \textbf{Working paper. Proceed with caution.}
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: references.bib
floatsintext: yes
numbersections: false
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
header-includes:
  - |
    \usepackage{float}
    \setlength{\parskip}{4pt}
output: 
  papaja::apa6_pdf:
    number_sections: true
    keep_tex: false
    highlight: kate
  papaja::apa6_docx: 
    number_sections: true
    keep_tex: false
---

```{r setup, include = FALSE}
# Packages
library(papaja)
library(janitor)
library(scales)
library(ellipse)
library(knitr)
library(latex2exp)
library(lme4)
library(ggpp)
library(patchwork)
library(latex2exp)
library(ggdist)
library(ggstance)
library(tidybayes)
library(distributional)
library(parameters)
library(brms)
library(tidyverse)

@@

# Plotting options
theme_set(
  theme_classic(base_size = 9) +
    theme(
      strip.text = element_text(color = "black", hjust = 0),
      panel.grid = element_blank(),
      line = element_line(linewidth = .25)
    )
)

# Model estimation options
dir.create("models", FALSE)
options(
  mc.cores = 4,
  brms.backend = ifelse(require(cmdstanr), "cmdstanr", "rstan")
)

```

@@

# Save variables as numerical & factor (might need for more complex contrast/eff/etc codes)
dat <- read_csv("data/heterogeneity_dataset1_traitvalence.csv") %>% 
  # Cleaning as in Bolger et al 2019
  filter(
    response.keys == "up",
    scale(rt) < 3,
    !(id %in% c(250, 257, 272))
  ) %>% 
  # Some more cleaning
  mutate(
    person = fct_anon(factor(id)),
    trial,
    valence = factor(valenceE, levels = c(-0.5, 0.5), labels = c("Negative", "Positive")),
    rt = logrt
  ) %>% 
  select(person, trial, valence, rt) %>% 
  arrange(person, trial)
```

@@

When building and testing theories of psychological phenomena, psychologists have long focused on asking whether an effect exists or not. Yet, establishing that an independent variable affects a dependent variable on average may not be a sufficient description of the phenomenon if the treatment effect is highly variable from one individual to another. The importance of such between-person variation, or *heterogeneity*, for theory development is widely recognized, yet rarely communicated sufficiently in the empirical literature [@bolgerCausalProcessesPsychology2019; @brandCausalEffectHeterogeneity2013].

One reason for the scarcity of reporting and interpreting heterogeneity is that psychologists still tend to analyze data with models that cannot address between-person variability, such as ANOVA [@bolgerCausalProcessesPsychology2019]. Although more appropriate mixed-effects (or multilevel, hierarchical) models are rapidly becoming more widespread, we believe a second important reason is that researchers do not yet have the conceptual and practical tools required to estimate and describe heterogeneity to a sufficiently informative degree. In the current work, we propose several practical methods, descriptions, and visualizations that allow researchers to better estimate, interpret, and report heterogeneity in a manner that maximizes empirical results' impact on theory building.

@@

Our aim is to outline methods for studying and communicating heterogeneity in a manner that fully takes uncertainty into account. We first introduce an example dataset from social psychology and review established procedures for estimating and communicating the expected heterogeneity in causal effects in the population. We then outline useful summaries of heterogeneity that focus on informative numerical and graphical descriptions, and note how they are limited by established methods' focus on point estimates of model parameters.

We then turn to our proposed method of interrogating and communicating causal effect heterogeneity using probabilistic methods. By estimating probability distributions over model parameters, these bayesian methods are uniquely suitable for examining heterogeneity and researchers' uncertainty about it. Moreover, advances in computational procedures and statistical software make these methods available to researchers with a minimal learning curve beyond established maximum likelihood based methods.

In the following sections, we detail calculations and graphics that make use of bayesian methods in communicating descriptions of heterogeneity such that uncertainty in the model parameters and, subsequently, heterogeneity, is taken into account. We then further extend the practice and consequent benefits of probabilistic inference for causal effect heterogeneity to comparing heterogeneity between populations and across timepoints in further example analyses.

@@

## Model 1

```{r}
#| label: tab-dat1
dat %>% 
  head() %>% 
  apa_table(
    span_text_columns = FALSE, 
    digits = c(0,0,0,2),
    caption = "First six rows of example dataset 1.", 
    note = "rt is log-transformed."
  )
```

In this section, we replicate @bolgerCausalProcessesPsychology2019's analysis, using their openly available data (https://github.com/kzee/heterogeneityproject). We first cleaned the data as in @bolgerCausalProcessesPsychology2019, leaving `r number(nrow(dat), big.mark = ",")` trials from 59 participants that were endorsed as self-relevant. (We present all our code in the online analysis supplement.) We show a sample of these data in Table \@ref(tab:tab-dat1). Then, we estimated the same statistical model with these data (Model 1; equations 1.1-1.3). First, we model the (log) reaction time of subject $j$ on trial $i$ as a random draw from a normal distribution with mean $\eta$ (*eta*, which can differ between trials $i$ and individuals $j$) and standard deviation $\sigma$ (*sigma*):

\begin{equation}
\tag{1.1}
\text{RT}_{ij} \sim \operatorname{Normal}(\eta_{ij}, \sigma^2).
\end{equation}

The lack of subscripts on $\sigma$ indicates that we assume it to be invariant across trials and subjects. Then, we specify a model of the mean $\eta_{ij}$ such that the regression coefficients capture our substantive questions:

\begin{equation}
\tag{1.2}
\eta_{ij} = \beta_0 + \gamma_{0j} + (\beta_1 + \gamma_{1j})\text{V}_{ij}.
\end{equation}

This equation includes two kinds of parameters: $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V), do not have subscripts. That means that they are not further modelled, and therefore sometimes called "fixed". We contrast coded valence such that negative words are coded as -0.5, and positive words as 0.5. This coding results in an intercept that corresponds to the average reaction time across negative and positive words, and a slope coefficient that reflects the difference in log(rt) between negative and positive words. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$ are subject-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the average reaction time for subject $j=1$. The $\gamma$ coefficients have subscripts, which in this context indicates that they are further modelled, and therefore sometimes called "random" [@gelmanDataAnalysisUsing2007]. As is standard in multilevel modelling, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

\begin{equation}
\tag{1.3}
\begin{bmatrix} 
\gamma_{0} \\ \gamma_{1}
\end{bmatrix} \sim 
\operatorname{MVN}(
\begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
\begin{bmatrix} 
\tau_{\gamma_{0}} &\rho_{\gamma_0\gamma_1} \\ 
\rho_{\gamma_0\gamma_1} &\tau_{\gamma_{1}} 
\end{bmatrix}
).
\end{equation}

In these equations, we assume that the subject-specific deviations have means of zero, standard deviations $\tau$ (*tau*), and a correlation $\rho$ (*rho*). Note that the $\gamma$s are zero-centered because the means are added to them in the above linear mixed-effects equation for $\eta_{ij}$. What these equations mean substantively is that the extent to which the effect of valence varies around the average effect ($\beta_1$) is estimated by the standard deviation $\tau_{\gamma_1}$. Moreover, $\rho$ indicates the extent to which individuals' average reaction times correlate with how much their reaction time is affected by valence.

With data described in Table \@ref(tab:tab-dat1) loaded to R, we can estimate this model using standard maximum likelihood methods as implemented in the R package lme4 [@batesFittingLinearMixedEffects2015; @rcoreteamLanguageEnvironmentStatistical2023].

@@

```{r fit1-brms}
fit1 <- brm(
  rt ~ 1 + valence + (1 + valence | person),
  data = dat,
  file = "models/brm-fit-1",
  control = list(adapt_delta = .95)
)
```

@@

tab1_lmer %>% 
  mutate(
    across(where(is.numeric), ~number(., .01)),
    MLE = str_glue("{Coefficient} [{CI_low}, {CI_high}]"),
  ) %>% 
  select(Parameter, `Estimate [95\\%CI]` = MLE) %>% 
  mutate(
    Parameter = m1_parnames
  ) %>% 
  apa_table(
    span_text_columns = FALSE,
    caption = "Parameter estimates from Model 1.",
    escape = FALSE
  )
mu1n <- tab1_lmer[2,2,T]
mu1 <- number(mu1n, .01)
sd1n <- tab1_lmer[4,2,T]
sd1 <- number(sd1n, .01)
```

We show a traditional summary of the estimated parameters from this model in Table \@ref(tab:tbl-fit-1). The expected effect of positive valence on RT for the average person is approximately `r mu1` log seconds, and the expected standard deviation describing this effect's degree of variability in the population is approximately `r sd1`.

## Heterogeneity distribution

```{r descriptors, include = FALSE}
# Heterogeneity interval
hi90_num <- mu1n + qnorm(c(0.05, .95)) * sd1n
hi90 <- paste(
  number(hi90_num, .01), 
  collapse = ", "
)
hi90_low <- paste(
  number(tab1_lmer[2,3,T] + qnorm(c(0.05, .95))*tab1_lmer[4,3,T], .01), 
  collapse = ", "
)

# Proportions
p_less_zero <- percent(pnorm(0, mu1n, sd1n), .1)
p_rope <- percent(pnorm(0.1, mu1n, sd1n) - pnorm(-0.1, mu1n, sd1n), .1)

# Ratio
ratio <- number(abs(sd1n / mu1n), .01)
```

```{r}
#| label: hdist-1
#| fig.height: 2.2
#| fig.width: 8
#| fig.env: "figure*"
#| fig.cap: The heterogeneity distribution and various descriptions of the expected heterogeneity in valence effects from Model 1. \textbf{A}. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. \textbf{B}. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. \textbf{C}. Proportion of negative valence effects in the population (dark). \textbf{D}. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

p0 <- ggplot() +
  aes(xdist = dist_normal(mu1n, sd1n)) +
  scale_x_continuous(breaks = extended_breaks()) +
  coord_cartesian(xlim = c(-0.55, 0.25)) +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(lineheight = rel(0.5))
    ) +
  scale_fill_manual(values = c("grey90", "grey80"), na.value = "white")
d1 <- dist_normal(mu1n, sd1n) %>% 
  mean_qi(.width = c(.66, .9))

p1 <- p0 +
  stat_slab(
    aes(fill = after_stat(level)), 
    .width = c(.68, .95), 
    p_limits = c(.000001, .999999), 
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$\\tau_1$"),
    x = mu1n + .05, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n + sd1n, y = .1, yend = .1,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed")
  ) +
  geom_text(
    label = TeX("$\\beta_1$"),
    x = mu1n-.05, y = .075,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n, y = .1, yend = 0,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed")
  ) +
  theme(axis.title.y = element_text(angle = 90))

p2 <- p0 + 
  stat_slab(
    aes(fill = after_stat(level)), 
    .width = c(.90, 1), 
    p_limits = c(.000001, .999999), 
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$HI_{90}$"),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = d1$.lower[2], xend = d1$.upper[2], y = .1, yend = .1,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed", ends = "both")
  )

p3 <- p0 + 
  stat_slab(
    aes(fill = after_stat(x < 0)), 
    p_limits = c(.000001, .999999), 
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX('$p^d$'),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  )

p4 <- p0 + 
  stat_slab(
    aes(fill = after_stat(between(x, -.1, .1))), 
    p_limits = c(.000001, .999999), 
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX('$p^{ROPE}$'),
    x = 0, y = .07,
    size = 3,
    parse = TRUE
  )

(p1 | p2 | p3 | p4) &
  scale_y_continuous(
    expand = expansion(c(0.0, 0.05))
  ) &
  labs(
    y = "Density",
    x = "Valence effect"
  ) &
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  ) &
  plot_annotation(tag_levels = "A")
```

The point estimates in Table \@ref(tab:tbl-fit-1) define the expected normal distribution of valence effects in the population, visualized in Figure \@ref(fig:hdist-1)A. As such, Normal(`r mu1`, `r sd1`^2^) is a valid description of heterogeneity in valence effects. However, it is an incomplete description of heterogeneity for two reasons. First, it does not necessarily communicate the degree of heterogeneity in clear and actionable terms. Second, it ignores the uncertainty with which the heterogeneity parameters are estimated.

### Interval descriptors

Having estimated the parameters of the heterogeneity distribution, we now turn to using them to define alternative, more informative descriptions of the degree of heterogeneity in valence effects in the population.

To begin with, and following @bolgerCausalProcessesPsychology2019, we used the point estimates in Table \@ref(tab:tbl-fit-1) to construct an expected *heterogeneity interval* that describes a range within which a certain percentage of the population's slopes are expected to fall. By convention, @bolgerCausalProcessesPsychology2019 and others have focused on the 95% heterogeneity interval ($HI_{95}$), but because there are already confusingly many quantities using the five percent cutoff, in this manuscript we focus on the 90% heterogeneity interval, and reserve 95% to describing uncertainties.

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use the .05 and .95 percentiles (which define the central 90% of the distribution). Then, we pass those percentiles and the estimated mean and standard deviation (see Table \@ref(tab:tbl-fit-1)) to the normal quantile function ($q$; `qnorm()` in R) to obtain the interval: $HI_{90} = q([.05, .95], \beta_1, \tau_1) = q([.05, .95],$ `r mu1`, `r sd1`) = [`r hi90`]. In words, this equation calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall within this interval. We illustrate this interval in Figure \@ref(fig:hdist-1)B.

### Proportion descriptors

For many applications, quantities such as proportions of slopes above or below some critical value, or within some critical range, might serve as more informative summaries. For example, we might ask "What proportion of individuals in the population respond faster to positively valenced words?" Thus, we first define a critical value, here zero. We then pass that value and the valence effect's estimated mean and standard deviation to the cumulative normal distribution function ($p$; `pnorm()` in R) to obtain the probability density in this distribution that falls below zero: $p^d = p(0, \beta_1, \tau_1) = p(0,$ `r mu1`, `r sd1`) = `r p_less_zero`. Specifically, this metric indicates the proportion of individuals in the population with negative effects. In words, we have estimated that `r p_less_zero` of the population respond faster to positively valenced words. We show this area under the curve in Figure \@ref(fig:hdist-1)C.

However, using zero as a critical value might not be sufficiently informative especially when theory allows defining a smallest effect sizes of interest, or what is known as a region of practical equivalence (to zero) [@anvariUsingAnchorbasedMethods2021; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014]. For example, we might know from theory that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the cumulative normal distribution function ($p$) to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = p(0.1, \beta_1, \tau_1) - p(-0.1, \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero (within the [-0.1, 0.1] interval). We visualize this area under the curve in Figure \@ref(fig:hdist-1)D.

### Ratio descriptors
